% §0 Abstract

Advanced Persistent Threats (APTs) unfold as multi-step causal chains---an initial compromise enables persistence, which enables lateral movement, culminating in data exfiltration. Yet provenance-based intrusion detection systems, which operate on graphs that explicitly encode causal dependencies, still detect attacks as coarse time windows, isolated entities, or graph snapshots. None identifies the actual causal event sequence that constitutes the attack narrative. We present the first APT detection system that operates at the granularity of causal event chains rather than time windows or individual entities. Our core contribution is a model-free Causal Coherence Metric ($\branchCoherence$) that quantifies the causal structural integrity of event chains extracted from provenance graphs, grounded in the operating system principle that processes are exclusive capabilities while files and sockets are shared resources. A calibration protocol based on elbow detection automatically discovers the natural causal horizon of an operating system from benign data alone, requiring no attack labels and no manual parameter tuning. Chains extracted under $\branchCoherence$ guidance are validated by a self-supervised autoregressive Transformer trained solely on benign event sequences via next-event prediction. On the DARPA Transparent Computing benchmark, our causal chains achieve [X]\% chain-level F1, outperforming window-based (EagleEye), random-walk-based (Sentient), and temporal-subgraph-based (GET-AID) alternatives by over 10\% absolute F1, while reducing investigation cost by more than 50$\times$ compared to the state-of-the-art window-level detector (KAIROS). The calibration protocol discovers distinct causal horizons across different operating system configurations, demonstrating automatic cross-platform adaptation.
