% Introduction
\section{Introduction}
\label{sec:introduction}

Modern enterprise and cloud environments consist of long-running hosts, service processes, and external connections that continuously generate large volumes of benign system activity. APT detection is necessary because advanced attackers do not always rely on a single conspicuous action. Instead, they progress through low-frequency, multi-stage operations that blend into normal administration and service behavior. The research challenge is therefore not merely to raise an anomaly alert, but to recover an interpretable attack process from noisy system behavior so that analysts can understand how the attack happened, which objects were affected, and how to respond.

Advanced Persistent Threats (APTs) typically unfold as multi-stage campaigns: an adversary first obtains an initial foothold, then performs persistence, privilege escalation, lateral movement, and data exfiltration~\cite{mitreatt}. Individual events in this process are often not rare. Launching a shell, reading a configuration file, opening a network connection, or writing to a temporary directory can all be legitimate operations on a production host. What separates an attack from benign activity is often not a single event, but whether a sequence of such events forms an interpretable causal chain.

System provenance graphs provide a natural data source for recovering such chains. Modern operating-system kernels mediate system calls and record dependencies among processes, files, and network objects. Organizing these audit records as a directed graph yields nodes for system entities and edges for kernel-mediated dependency events. Compared with application logs or aggregate statistics, provenance graphs have two important advantages: they provide broad coverage of kernel-mediated system interactions; and unless the attacker already has kernel-level capability, the corresponding audit records are difficult to tamper with.

Directly following dependencies in a provenance graph, however, leads to \emph{dependency explosion}~\cite{kingbacktracking}: expanding forward or backward from a suspicious event can quickly include large amounts of unrelated benign activity. Existing provenance-based intrusion detection systems (PIDS) mostly detect anomalies at the graph, window, entity, or edge level~\cite{unicorn2020,yang2023prographer,cheng2024kairos,jia2024magic,wang2022threatrace,edgetrace2025}, leaving attack-chain reconstruction to downstream post-processing or manual investigation. Recent Transformer-based methods over provenance data typically construct sequences using fixed time windows, random walks, or temporal subgraphs~\cite{eagleeye2024,sentient2026,getaid2025}; these sequences need not respect provenance-edge direction or entity-identity continuity.

The lack of explicit causality has three direct consequences. First, temporally adjacent events can be treated as related even when they are only concurrent benign activities, causing models to learn incidental correlations. Second, point or local anomaly alerts do not naturally form a complete attack narrative, so analysts must still decide which events actually influenced one another. Third, reconstructing paths after detection using graph reachability often reintroduces many irrelevant dependencies, making explanations long, noisy, and hard to verify. In short, without causal constraints, a detector may know that ``something is abnormal'' but still fail to explain ``which events together constitute the attack.''

Our research idea is to move causality into the construction of the detection object. We first ask which events can plausibly influence one another under system semantics, thereby forming a structurally credible causal chain, and only then ask whether that chain deviates from benign behavior. This design has three intuitive benefits. It restricts the search space to events with dependency support, mitigating dependency explosion; it preserves the continuous narrative from foothold to later actions, reducing manual stitching; and it separates ``whether a chain is valid'' from ``whether a valid chain is anomalous,'' making detection results easier to explain and evaluate. Chain-level APT detection therefore requires an upstream definition: \emph{which event sequences should count as structurally valid causal chains?}

Causal chains alone, however, do not complete attack detection. A structurally valid chain can come from either normal administration or a real attack; the difference often lies not in a single event type, but in the ordering, context, and staged semantics of events along the chain. Relying on hand-written rules or known attack templates would limit the ability to detect unknown APTs. We therefore further train a self-supervised Transformer detector on benign causal chains, so that the model learns normal chain-level behavior and flags chains that significantly deviate from it. In this design, causal-chain extraction ensures that the detection object is a meaningful behavioral process, while the Transformer detector decides whether that process is anomalous; together they form chain-level APT detection.

This paper answers the above questions. We first give the basic definitions of system provenance graphs and chain-level detection objects in the method section. We then define a chain coherence score $\branchCoherence$ that requires neither attack labels nor a trained model. The score is not a maliciousness score and is not interpreted as a probability; it measures whether adjacent chain links are supported by observable dependency evidence in the audit log. We then build a chain-level detection pipeline: an edge-level seed scorer selects events worth investigating, $\branchCoherence$ guides bidirectional causal-chain extraction, and a self-supervised Transformer detector trained on benign chains produces the anomaly score $A(\chain)$.

Our contributions follow a three-layer structure---problem formulation, chain extraction, and end-to-end validation:
\begin{enumerate}
\item \textbf{We reformulate APT detection as causal-chain recovery.} Unlike window-, entity-, or edge-level alerting, our work directly treats the event chain needed for investigation as the detection object. We separate two questions: whether a chain is structurally valid on the provenance graph, and whether a structurally valid chain deviates from benign behavior. The former is bounded by kernel-audit semantics; the latter is learned from the benign distribution. This decomposition explains why chain-level detection is well-founded: both benign and attack chains must obey system dependency structure, but attack chains are semantically anomalous.
\item \textbf{We propose a method for extracting trustworthy causal chains from large provenance graphs.} To address dependency explosion, we do not blindly expand graph reachability. Instead, $\branchCoherence$ scores whether adjacent events have sufficient dependency support, and uses that score to rank, prune, and truncate candidate chains. The score depends only on observable audit-log evidence---including process identity continuity, whether shared resources are perturbed by third parties, and the finite identity scope of system entities---so it can be calibrated without attack labels or attack templates.
\item \textbf{We build and evaluate a zero-attack-prior chain-level APT detection system.} The system first extracts structurally credible candidate chains, then uses a self-supervised sequence model to detect semantic anomalies, and evaluates outputs with chain-level metrics for attack relevance and complete narrative recovery. On DARPA Transparent Computing data, we report end-to-end detection, ablation, cross-engagement generalization, and efficiency results, showing that chain-level outputs can reduce analyst investigation cost more directly than traditional window or node alerts.
\end{enumerate}

The remainder of this paper is organized as follows. \S\ref{sec:method} introduces system provenance graphs and describes causal-chain extraction and sequence modeling. \S\ref{sec:setup} gives the experimental setup. \S\ref{sec:evaluation} reports results. \S\ref{sec:discussion} discusses scope and limitations. \S\ref{sec:related} compares related work. \S\ref{sec:conclusion} concludes.
