% Background and Problem Formulation
\section{Background and Problem Formulation}
\label{sec:background}

\subsection{System Provenance Graphs}

A provenance graph $\graph=(V,E)$ is a directed graph constructed from kernel-level audit logs. Each node $v\in V$ represents a system entity, such as a process, file, or network connection. Each edge $e=(u\rightarrow v,t,\tau)\in E$ represents a kernel-mediated dependency event, where entity $u$ affects entity $v$ at time $t$ through event type $\tau$. Typical events include process creation, program execution, file reads and writes, and network sends and receives.

Provenance graphs are useful for APT detection because they preserve the dependency structure among system entities. Rather than asking only whether a time window or an entity is anomalous, provenance allows an analyst to ask a more investigation-oriented question: what caused a suspicious behavior, and what did it subsequently cause? The challenge is scale. The DARPA Transparent Computing E3 CADETS trace contains hundreds of thousands of nodes and nearly thirty million edges; unconstrained forward or backward tracing quickly absorbs unrelated benign activity.

\subsection{Chain-Level APT Detection}
\label{sec:problem}

We define chain-level APT detection as follows. Given a provenance graph $\graph$ and a set of suspicious seed events, the detector outputs time-ordered event chains $\chain=[e_1,\ldots,e_n]$ that satisfy two requirements. First, the chain should be \emph{dependency-plausible}: adjacent events should be connected by audit-supported dependency evidence, not merely by temporal proximity or graph reachability. Second, the chain should be semantically unusual relative to benign system behavior, and thus worthy of analyst investigation.

This definition deliberately separates two tasks. \emph{Chain extraction} recovers structurally credible candidate chains; \emph{anomaly detection} determines whether those candidates deviate from the benign distribution. The former does not require attack labels and should not directly express maliciousness. The latter operates on the candidate chains produced by the former.

\subsection{Audit Model and Assumptions}
\label{sec:audit_model}

All claims about dependency evidence in this paper are scoped to the following audit model.

\paragraph{Trusted computing base.} We trust the operating-system kernel, its system-call mediation path, the audit subsystem, the audit-event collector, and persistent storage of audit records. User-space processes, application logs, and container user-space utilities are outside the TCB. Kernel compromise, audit-collector shutdown, and targeted audit-buffer overflow that silently drops events are out of scope.

\paragraph{Event coverage.} Table~\ref{tab:event_semantics} lists the main CADETS audit edge types used in this paper. Events outside the table are not interpreted as observable dependencies; in particular, shared memory, some \texttt{mmap}-based IPC, and signal-based IPC are out of model.

\begin{table*}[t]
\centering
\caption{Audit-event semantics used in this work.}
\label{tab:event_semantics}
\small
\begin{tabular}{llllll}
\toprule
Event type & Source & Target & Time semantic & Direction & Note \\
\midrule
EVENT\_EXECUTE & process & binary file & exec start & control/data & program image switch \\
EVENT\_READ & file & process & syscall return & data in & no content or offset \\
EVENT\_WRITE & process & file & syscall start & data out & append/offset may be missing \\
EVENT\_FORK & parent process & child process & fork return & control/identity & child inherits parent state \\
EVENT\_SENDTO & process & netflow & syscall start & data out & flow identifier only \\
EVENT\_RECVFROM & netflow & process & syscall return & data in & flow identifier only \\
EVENT\_OPEN & process & file & syscall return & state & resource access \\
EVENT\_CLOSE & process & file & syscall return & state & resource release \\
\bottomrule
\end{tabular}
\end{table*}

\paragraph{Entity identity.} Process identity is represented by $(pid,\text{creation\_time})$ to avoid PID reuse; crossing an \texttt{execve} boundary preserves the PID tuple but weakens identity continuity. File identity is represented by inode and mount identifier, with rename tracking; hard links are treated as the same entity. Socket identity uses the connection 5-tuple and audit object identifier. When thread identifiers are available, cross-thread links are explicitly down-weighted.

\paragraph{Missing events.}
\label{sec:audit_missing}
We do not infer the content of missing events. If an event class is outside the audit model, or if an expected edge is missing inside a chain, the coherence score uses a smoothing floor $\phi_{\min}$ to avoid numerical collapse and a long-gap threshold $\tau_{\text{gap}}$ to prevent long silent intervals from being interpreted as valid dependency links.

\paragraph{Out-of-model IPC.}
\label{sec:audit_shm}
Shared memory, implicit communication through \texttt{mmap}, signal-based IPC, and some anonymous-pipe cases lack sufficient semantics in the default CADETS event set. Attacks relying primarily on these channels may be missed; we discuss this scope limitation in \S\ref{sec:discussion}.

\subsection{Research Challenges}

Chain-level detection faces three challenges. First, graph reachability over-approximates dependency causality: two events may be connected through a shared resource, but intervening third-party writes can break the actual dependency. Second, both attack chains and benign chains obey operating-system dependency structure, so structural coherence alone cannot imply maliciousness. Third, chain-level outputs require their own evaluation protocol, because conventional window- or node-level metrics do not measure whether an attack narrative has been recovered.

The rest of the paper addresses these challenges. \S\ref{sec:method} tightens candidate chains using audit-supported coherence. \S\ref{sec:setup} defines the chain-level evaluation protocol. \S\ref{sec:evaluation} evaluates structural coherence, end-to-end detection, and complete narrative recovery.
