% Experimental Setup
\section{Experimental Setup}
\label{sec:setup}

This section describes datasets, baselines, evaluation metrics, and implementation details. Results are reported in \S\ref{sec:evaluation}.

\subsection{Datasets and Splits}

We use the DARPA Transparent Computing dataset. The main experiments use the E3 CADETS engagement, and cross-engagement experiments use E5 THEIA/Trace. The E3 trace contains approximately 268{,}242 nodes and 29{,}727{,}441 edges over five days of system activity. Following prior PIDS work, days 2--4 are treated as benign: days 2--3 are used for parameter calibration and model training, and day 4 is held out as benign validation data. Day 6 contains the ground-truth attack and is used for testing. Attack labels follow the DARPA TC annotations used by MAGIC and ThreaTrace.

\subsection{Baselines}

We compare against baselines that cover different detection granularities and chain-construction strategies:
\begin{itemize}
\item \textbf{Window-level methods:} KAIROS~\cite{cheng2024kairos} and an EagleEye-style fixed-window sequence baseline.
\item \textbf{Entity/edge-level methods:} MAGIC~\cite{jia2024magic}, ThreaTrace~\cite{wang2022threatrace}, and an EdgeTrace-style edge scorer~\cite{edgetrace2025}.
\item \textbf{Sequence/subgraph methods:} Sentient-style random-walk segmentation~\cite{sentient2026} and GET-AID-style temporal-subgraph segmentation~\cite{getaid2025}.
\item \textbf{Path-reconstruction methods:} SPARSE/Holmes-style rule tracing~\cite{sparse2024,holmes2019} and SLOT-style post-hoc chain reconstruction~\cite{slot2025}.
\end{itemize}

For baselines that do not directly output chains, we construct candidate chains around their flagged windows, nodes, or edges using 1--2 hop directed BFS, and choose the setting that is more favorable to the baseline. This adaptation biases the comparison toward baselines; any remaining gap therefore reflects chain-construction quality rather than the matching convention.

\subsection{Evaluation Metrics}
\label{sec:eval:metrics}

We report results at three levels.
\begin{itemize}
\item \textbf{Dependency-plausible chains (DPC).} This level evaluates whether a chain is structurally supported by audit evidence. Metrics include the discriminative power of $\log\branchCoherence$ between benign and random chains, degradation under negative-control perturbations, and resource-survival goodness-of-fit.
\item \textbf{Attack-relevant chains (ARC).} This level evaluates overlap between predicted chains and attack ground truth. The primary metric is Edge-IoU: for each predicted chain $C_p$, we compute $\max_g \mathrm{IoU}(E(C_p),E(C_g))$. We also report attack-node recall and temporal overlap.
\item \textbf{Complete attack narratives (CAN).} This level evaluates how many top-ranked chains an analyst must inspect to cover attack stages such as entry, persistence, lateral movement, and exfiltration. This metric directly captures investigation cost.
\end{itemize}

In addition to chain-level metrics, we report Top-$K$ results for $K\in\{10,50,100\}$, and measure per-component runtime, throughput, and memory use.

\subsection{Implementation Details}

The current implementation uses the pretrained TGN reconstruction loss from KAIROS as the edge-level seed score. Edges whose loss exceeds the 95th percentile of the benign distribution are used as seeds. Chain-extraction parameters $\hoplimit^*$, $\timedecay^*$, $\graphdecay^*$, $\branchpenalty^*$, $\phi_{\min}$, $\tau_{\text{gap}}$, and $\beta_p$ are calibrated only on benign data.

The sequence model is a 3-layer Transformer encoder with hidden dimension 128 and 4 attention heads, trained autoregressively for next-event prediction. Training chains are generated from a 50\%/50\% mixture of uniform random anchors and benign high-loss anchors, balancing broad benign coverage with online covariate matching. The final anomaly score $A(\chain)$ is calibrated by empirical CDFs of loss components on the benign validation set; the alert threshold is the 99.9th percentile of the benign validation distribution.
