% Results
\section{Results}
\label{sec:evaluation}

This section reports results around four research questions: whether the chain coherence score is effective (RQ1), whether chain-level detection improves APT investigation (RQ2), how much each component contributes (RQ3), and whether the method generalizes and runs efficiently (RQ4).

\subsection{RQ1: Is the Chain Coherence Score Effective?}
\label{sec:metric}

RQ1 evaluates $\branchCoherence$ itself, without training or using the sequence model. The goal is to verify that the score measures audit-supported structural coherence rather than graph density, temporal proximity, or random reachability.

\paragraph{Benign-vs-random chain discrimination.} We extract chains from 100 TGN seeds on benign day 4, and compare them with random BFS chains constructed without direction or identity constraints.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S5\_BENIGN\_RANDOM --- AUC and histogram for $\log\branchCoherence$ distinguishing benign and random chains. -->}\end{flushleft}
Benign chains are expected to have higher $\log\branchCoherence$, while random chains should degrade toward the $\phi_{\min}$ floor due to direction violations, identity discontinuity, and long time gaps.

\paragraph{Negative-control validation.} We construct four perturbations: time shuffling, direction reversal, entity randomization, and resource-interleaving shuffling. Each preserves some surface statistics while breaking one audit-supported factor.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S5\_NEGCTRL --- median $\log\branchCoherence$, IQR, and gap to benign median under four perturbations. -->}\end{flushleft}
If all four perturbations significantly reduce the score, the metric is sensitive to temporal order, edge direction, entity identity, and resource-survival semantics.

\paragraph{Resource-survival goodness-of-fit.} For regular files, we report KS statistics and $\chi^2$ goodness-of-fit for the Poisson baseline; if the fit fails, we use a Kaplan--Meier empirical survival function estimated from benign data. For pipes, we report FIFO-violation rate. For sockets, we report the fraction of send/recv pairs lying within the same connection lifetime. For shared memory and ambiguous IPC, we report prevalence only because default audit logs lack stable dependency semantics for them.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S5\_SURVIVAL --- per-resource-class fit statistics, edge fraction, and out-of-model event prevalence. -->}\end{flushleft}

\paragraph{Long-lived-process false-link rate and sensitivity.} On benign day 4, we identify daemons with lifetime longer than one hour and sample intra-process event pairs at different $\Delta t$, reporting the fraction with $\phi_{\text{proc}}>0.5$. We also report sensitivity to $\phi_{\min}$, $\tau_{\text{gap}}$, and $\beta_p$.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S5\_DAEMON / GAP\_S5\_SENSITIVITY --- daemon false-link rate and three hyperparameter sensitivity tables. -->}\end{flushleft}

\subsection{RQ2: Does Chain-Level Detection Improve APT Investigation?}

\paragraph{Main detection results.} We train the sequence model on benign chains from days 2--4 and test on the day-6 attack. All sequence-based baselines use the same Transformer architecture and training procedure; only the chain or segment construction strategy changes, isolating the contribution of causal-chain extraction.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S6\_MAIN --- Edge-IoU, attack-node recall, and investigation cost across all methods at Top-$K\in\{10,50,100\}$. -->}\end{flushleft}
We expect the proposed method to achieve higher chain-level Edge-IoU and node recall than window-level, entity-level, and post-hoc reconstruction baselines, while reducing the number of edges/nodes analysts must inspect to recover a complete attack narrative.

\paragraph{Seed-recall decomposition.} To avoid conflating seed-scorer quality with chain-extractor quality, we report
\[
\mathrm{Recall}_{\text{end-to-end}} = \mathrm{Recall}_{\text{seed}} \times \mathrm{Recall}_{\text{chain}\mid\text{seed-covered}}.
\]
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S6\_SEEDRECALL --- seed recall, conditional chain-layer recall, and their product. -->}\end{flushleft}

\paragraph{Attack narrative recovery.} On the nginx exploit case, we label ground-truth attack edges by stage---entry, persistence, lateral movement, and exfiltration---and measure how many top-ranked predicted chains must be merged to cover all stages.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S6\_CAN --- stage-coverage curve as a function of merged-chain count, compared with KAIROS/MAGIC baselines. -->}\end{flushleft}

\subsection{RQ3: How Much Does Each Component Contribute?}

We ablate three groups of components. Extractor ablations include removing causal direction enforcement, replacing the mean-log score with an arithmetic mean, removing branch-weighted refinement, and disabling $g_{\text{lifetime}}$ and $\beta_p$ inside $\phi_{\text{proc}}$. Feature/model ablations remove global node statistics, remove sinusoidal time encoding, or replace the Transformer with an LSTM or MLP. Anchor-sampling ablations compare C-random-only, C-seed-only, and the default C-mixture setting.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S6\_ABLATION --- ablation bar charts and per-item metric table. -->}\end{flushleft}

\subsection{RQ4: Does the Method Generalize and Run Efficiently?}

\paragraph{Cross-engagement generalization.} We apply the same protocol to DARPA E5. Calibration, training, and testing remain unchanged; only the parameters are rediscovered on E5 benign data.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S6\_CROSS --- E3/E5 calibration curves, discovered parameters, and chain-level metrics. -->}\end{flushleft}
We expect different engagements to produce different $\hoplimit^*$, $\tau_{\text{gap}}$, and resource-survival parameters, while maintaining stable chain-level performance.

\paragraph{Efficiency.} In the current implementation, processing one E3 day with about $5$M events is dominated by seed scoring. Chain extraction is an incremental CPU-side cost, and sequence-model inference after extraction is nearly immediate.
\begin{flushleft}\small\texttt{<!-- DATA\_NEEDED: GAP\_S6\_EFFICIENCY --- per-component latency, memory footprint, and throughput comparison. -->}\end{flushleft}
