% Discussion
\section{Discussion}
\label{sec:discussion}

\subsection{Scope and Deployment Model}

The proposed system is designed for asynchronous out-of-band detection rather than inline blocking. It processes audit events in a sliding graph buffer: the edge-level seed scorer first identifies events worth investigating, and chain extraction plus sequence-model inference runs only when a seed is triggered. This design is appropriate for APT investigation, where campaigns unfold over hours or days; it trades millisecond-level blocking for higher throughput and interpretable chain-level outputs.

Our evaluation uses per-day batch processing to remain comparable with prior PIDS work. In deployment, the same pipeline can be implemented as micro-batch streaming. As long as audit events enter the graph buffer in temporal order, the chain extractor and eCDF calibration mechanism remain unchanged.

\subsection{Limitations}

\textbf{$\branchCoherence$ is not a maliciousness score.} It measures whether a chain has audit-supported structural coherence, not whether the chain is malicious. Both benign and attack chains can be highly coherent; maliciousness is determined by the sequence model's anomaly score $A(\chain)$. The negative-control and sensitivity experiments in \S\ref{sec:metric} validate that the score captures structural properties, but we do not interpret it as a calibrated probability.

\textbf{Recall is bounded by the seed scorer.} Chain extraction is triggered by seed edges. If an attack step is never surfaced by the seed scorer, the downstream extractor cannot recover it. We therefore report seed recall and chain-layer conditional recall separately, avoiding attribution of seed-scorer quality to the chain extractor.

\textbf{Path representation limits branched campaigns.} The current scoring object is a seed-centric linearized path; multiple alerts are merged into an analyst-facing investigation graph only after scoring. This is suitable for sparse, multi-stage APTs, but less natural for ransomware, worms, or mass scanning. Native DAG scoring is a direct extension.

\textbf{Out-of-model communication channels.} Shared memory, some \texttt{mmap}-based IPC, and signal-based IPC are outside the default audit model. Attacks relying primarily on these channels may be missed; handling them requires stronger audit instrumentation or additional runtime monitoring.

\textbf{Benign training assumption.} We assume the training data is attack-free, which is standard in provenance anomaly detection. In real deployments, this assumption can be relaxed with robust statistics, coarse contamination filtering, or analyst-confirmed benign windows.

\textbf{Cross-engagement, not full cross-platform generalization.} DARPA E3 and E5 come from the same program. Although they exercise different scenarios, their audit infrastructure is similar. Our results therefore establish cross-engagement adaptation, not full cross-OS or cross-audit-framework generalization.

\subsection{Future Work}

Future work includes four directions. First, extending the current batch implementation to live audit-log streaming. Second, designing dependency-causality-aware seed scorers to reduce reliance on a generic TGN reconstruction loss. Third, extending $\branchCoherence$ from linear paths to native DAG scoring, which would better capture branched attacks. Fourth, integrating multi-host provenance; in that setting, dependencies are no longer fully mediated by a single host kernel, and the audit model must be extended accordingly.
