% Conclusion
\section{Conclusion}
\label{sec:conclusion}

The central claim of this paper is that APT detection should not only answer where or when something anomalous happened; it should also answer which events together constitute the attack. System provenance graphs make this goal possible because kernel audit records preserve dependencies among processes, files, and network objects. Yet raw reachability is too broad: direct tracing leads to dependency explosion.

Around this problem, the paper makes three layered contributions. First, it formulates APT detection as chain-level recovery and separates structural validity from semantic anomaly. Second, it proposes a $\branchCoherence$-guided causal-chain extraction method that uses observable system-dependency evidence to select structurally credible candidate chains from large provenance graphs. Third, it builds a self-supervised detection system that requires no attack labels or attack templates, and evaluates the recovery of attack-relevant chains and complete attack narratives with chain-level metrics.

The significance of this design is that it moves the analyst's desired investigation object into the detector itself. The output is no longer merely a suspicious window or suspicious node, but a candidate attack chain organized by causal relationships. Our results show that chain-level modeling can support APT investigation more directly and reduce downstream manual reconstruction cost while preserving a zero-attack-prior setting.
